Program Transformations for miniKanren

### 2020-07-07 papers/aplas/text/intro.tex
Edits to intro

The core feature of relational programming is the ability to run a program in various directions by executing goals with free variables.
The distribution of free variable occurrences determines the direction of relational search.
Having specified a relation for adding two numbers, one can also compute the subtraction of two numbers or find all pairs of numbers which can be summed up to get the given one.
Program synthesis can be done by running \emph{backwards} a relational interpreter for some language.
In general, it is possible to create a solver for a recognizer by translating it into \mk and running in the appropriate direction~\cite{lozov2019relational}.

@@

Specialization or partial evaluation~\cite{jonesbook} is a technique aimed at improving the performance of a program given some information about it beforehand.
It may either be a known value of some argument, its structure (i.e. the length of an input list) or, in case of a relational program, --- the direction in which it is intended to be run.
An earlier paper~\cite{lozov2019relational} showed that \emph{conjunctive partial deduction}~\cite{de1999conjunctive} can sometimes improve the performance of \mk programs.
Unfortunately, it may also not affect the execution time of a program or even make it slower.

Control issues in partial deduction of logic programming language \pro have been studied before~\cite{leuschel2002logic}.
The ideas described there are aimed at left-to-right evaluation strategy of \pro.
Since the search in \mk is complete, it is safe to reorder some relation calls within the goal for better performance.
While sometimes conjunctive partial deduction gives great performance boost, sometimes it does not behave as well as it could have.

In this paper, we show on examples some issues which conjunctive partial deduction faces.
We also describe a novel approach to partial deduction of a relational programming language \mk.
We compare it to the existing specialization system (\ecce) on several programs and discuss why some \mk programs run slower after specialization.

>>>>

The core feature of relational programming is the ability to run a program in various directions by executing goals with free variables.
The distribution of free variable occurrences determines the direction of relational search.
For example, having specified a relation for adding two numbers, one can also compute the subtraction of two numbers or find all pairs of numbers which can be summed up to get the given one.
Program synthesis can be done by running a relational interpreter for some language \emph{backwards}.
In general, it is possible to create a solver for a recognizer by translating it into \mk and running in the appropriate direction~\cite{lozov2019relational}.

@@

Specialization or partial evaluation~\cite{jonesbook} is a technique aimed at improving the performance of a program given some information about it beforehand.
It may either be a known value of some argument, its structure (i.e. the length of an input list) or, in case of a relational program, --- the direction in which it is intended to be run.
An earlier paper~\cite{lozov2019relational} has shown that \emph{conjunctive partial deduction}~\cite{de1999conjunctive} can sometimes improve the performance of \mk programs.
Unfortunately, it may also not affect the execution time of a program or even make it slower.

Control issues in partial deduction of logic programming language \pro have been studied before~\cite{leuschel2002logic}.
Unlike Prolog, where atoms in the rhs of a clause cannot be arbitrarily reordered without changing the semantics of a program, in miniKanren the subgoals of conjunction/disjunction can be freely switched.
This opens a yet another possibility for optimization, not taken into account by approaches initially developed in the context of conventional logic programming.
% The ideas described there make use of the deterministic search strategy of \pro.
% However in \mk it is allowed to reorder some relation calls within the goal for better performance.
% While sometimes conjunctive partial deduction gives great performance boost, sometimes it does not behave as well as it could have.

In this paper, we study issues which conjunctive partial deduction faces being applied for \mk.
We also describe a novel approach to partial deduction for relational programming, \emph{conservative partial deduction}.
We implement this approach and compare it with the existing specialization system (\ecce) on several programs and discuss why some \mk programs run slower after specialization.
